---
title: "A gentle introduction to mathematical simulation in R"
format: html
---
## Required packages
```{r}
library(deSolve) #Differential equation solver
library(reshape2) #Data.frame formatter
library(ggplot2)
library(gganimate)
```

# 1. How does a population grow?
## 1A.1: Writing a function to generalize population model

```{r}
pop_Growth <- function(){
  ______
  ______
}
```

## 1A.2: Simulating and seeing the output

```{r}
head(______, )
```

## 1B.1: Creating a grid of parameter options

```{r}
param_grid <- expand.grid(______)

param_grid
```

## 1B.2: Multiple simulations

```{r}
results <- do.call(
  rbind,
  lapply(______))
```

## 1B.3: Plotting it out

```{r}
results_subset <- subset(results, _______ & ________)
```

```{r}
ggplot(results_subset, aes(x = time_step, y = N, group = as.factor(N0), color = as.factor(N0))) + 
  geom_point() + 
  geom_line()+ 
  xlab("Time (days)") + 
  ylab("Abundance") + 
  labs(color = "N0") +
  theme_classic() + 
  theme(axis.text = element_text(size = 13),
       axis.title = element_text(size = 14))
```


# 2. LOTKA-VOLTERRA PREDATOR PREY MODEL (MODEL)

## 2A.1: Creating the deSolve function

```{r}
LV_prey_pred <- function(t, state, param) {
  with(as.list(c(state, param)), {
  ______
  ______
 return(list(c(V = dV, P = dP)))
  })
}

```

## 2A.2: Choosing the initial conditions

```{r}
times <- _____________
param <- c(r= 1, k = 0.02, a = 0.5, mu = 0.2)
states <- c(V = 1 ,P = 2)
```

## 2A.3: Running the simulation
```{r}
lv_results <- ode(y =_____________, times = _____________, func =_____________ , parms = _____________)

head(lv_results, 10)
```

## 2A.4: Running the simulation

```{r}
lv_results <- ode(y =_____________, times = _____________, func =_____________ , parms = _____________)

head(lv_results, 10)
```

## 2A.5: Catching the error

```{r}
LV_prey_pred_mistake <- function(t, state, param) {
  with(as.list(c(state, param)), {
    
    dV <- (r * V) + (k * V * P) # The number of prey increases... with predation
    dP <- (a * k * V * P) -  mu * P

    return(list(c(V = dV, P = dP)))
  })
}
```

```{r}
param_no_b <- param
param_no_b["____"] <- ____
```

```{r}
lv_test_no_b <- data.frame(ode(y = ______, times = ______, func = ______, parms = ______))
```

```{r}
plot(lv_test_no_b[,_____], type = 'l',xlab = "Time", ylab = "Abundance of prey")
```

# 3. LOTKA-VOLTERRA PREDATOR PREY MODEL (VISUALIZING)

## 3A.1: Melting the output

```{r}
lv_results_melt <- melt(______, id.vars = ______)
```

## 3A.2: Plotting the output (static)

```{r}
ggplot(______, aes(x = ______, y = ______, color =______, group = ______)) + 
    geom_line() +
     theme_classic() +
     xlab("Time (Day)") + 
     ylab("Abundance") +
    theme(axis.text = element_text(size = 12))
```

## 3A.3: Plotting the output (dynamic)

```{r}
animation <- ggplot(______, aes(x = ______, y = ______, color =______, group = ______)) + 

  geom_line() +
  theme_classic() +
  xlab("Time (Day)") +
  ylab("Abundance") +
  transition_reveal(______) +
  theme(axis.text = element_text(size = 12))

animate(animation, duration = 15) 

```

# 4. LOTKA-VOLTERRA PREDATOR PREY MODEL (PHASEPLOT)
## 4A.1: Creating a function to calculate nullclines


```{r}
calculate_nullclines <- function(param_df){
 v_nullcline <- param_df["mu"]/_____
 p_nullcline <- _____/param_df["k"]

 return(data.frame(V_null = v_nullcline, P_null = p_nullcline))
}

null_cline_df <- calculate_nullclines(param)
```

## 4A.2: Plotting the nullclines

```{r}
basis_gg<- ggplot() + 
  geom_vline(xintercept  =______________________, color = 'red') + 
  geom_hline(yintercept = ______________________, color = 'blue') + 
  annotate("point", x = null_cline_df[,"V_null"], y = null_cline_df[,"P_null"] ) +
  xlab("______________________") + 
  ylab("______________________") + 
  theme_classic()
```

## 4A.3: Plotting the trajectory

```{r}
basis_gg + 
  geom_path(data = lv_results_____________)
```

## 4A.4: Plotting the trajectory (dynamic)


```{r}
animation_2<- basis_gg + 
  geom_path(data = lv_results,aes(x = V, y = P )) + 
  geom_point(data = lv_results,aes(x = V, y = P ), size = 3) + 
  __________________
  theme(axis.text = element_text(size = 12))

animate(animation_2, duration = 15) 
```


# 5. LOTKA-VOLTERRA PREDATOR PREY MODEL (ADVANCE)
## 5A.1: Adding constraint to our prey's growth



```{r}
LV_prey_pred_C <- function(t, state, param) {
  with(as.list(c(state, param)), {
    
    dV <- _____________ - (b * V * P)
    dP <- (a * b * V * P) -  mu * P

    return(list(c(V = dV, P = dP)))
  })
}

times <- seq(1,365, 0.1) # 5 years
param_C <- c(r= 0.25, b = 0.05, a = 0.5, mu = 0.2, C = 20)
states <- c(V = 1,P = 2)

lv_results_C <- data.frame(ode(y = states, times = times, func = LV_prey_pred_C, parms = param_C))

```

## 5A.2: Visualizing dynamics (static)

```{r}
lv_results_C_melt <- melt(lv_results_C, id.vars = 'time')
```

So let's plot it out:

```{r}
ggplot(lv_results_C_melt, aes(x = time, y = value, color = variable, group = variable)) + 
    geom_line() +
     theme_classic() +
     xlab("Time (Day)") + 
     ylab("Abundance") +
    theme(axis.text = element_text(size = 12))
```

## 5A.3: Visualizing dynamics (dynamic)

```{r}
animation_3 <- ggplot(lv_results_C, aes(x = V, y = P)) +
    xlab("Prey") + 
    ylab("Predator") + 
    geom_path() + 
    theme_classic() + 
    transition_reveal(time) 

animate(animation_3, duration = 15) 

```
